---
title: "scShardSplitRef"
author: "Irina Kuznetsova"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scShardSplitRef}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(scShardSplitRef)
```


## 1. Package overview  

sc_ShardSplitRef designed for researchers working with single-cell data from species that require building a custom reference genome, particularly when one or more contigs in the reference FASTA exceed the Cell Ranger limit of 536.8 Mb (2²? bases). sc_ShardSplitRef automates this splitting process, generating contig fragments that comply with Cell Ranger requirements. The software outputs FASTA and GTF files containing the split contigs, which can be directly processed by cellranger-arc mkref.  
This real-world example demonstrates a step-by-step workflow for downloading and preparing barley (MOREXv3) reference files (GTF and FASTA) for cellranger-arc mkref.  




## 2. Installation  
Start by installing the BiocManager package:  

#```{r}
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("scShardSplitRef")
#```  


Then, install the package(s) you would like to use:  

#```{r}
#BiocManager::install(c("dplyr", "stringr", "utils"))
#``` 



## 3. Accessing Package Help and Vignettes  


```{r}
library(scShardSplitRef)
help(package="scShardSplitRef")
vignette(package="scShardSplitRef")
vignette(package="scShardSplitRef", "scShardSplitRefHOWTOs")
?scShardSplitRef
```  


## 4. Standard workflow
### 4.1 Quick start

## **RUN THE ANALYSIS BY CALLING THE FUNCTIONS**
##### 1) Check if gene features are overlapping with the centromere position  

```
gtf_path_T="/toy_input/A1_toy_all_no_overlap.gtf"
centromere_path_T="/toy_input/IN0_toy_centromeres_for_gtf.bed"
out_path_T="/toy_output/"
centromere_split = prepare_centromere_split(
    gtf_path_T, 
    centromere_path_T, 
    out_path_T
    )
```

```
No overlapping gene features detected — use the original centromere coordinates BED file.
Use:
/toy_input/IN0_toy_centromeres_for_gtf.bed as the input to modify GTF file.
and:
/toy_output/IN0_original_centromeres_for_FASTA.bed for splitting FASTA
```

###### Depending on the input data, the message could be:  

```Overlapping gene features with centromere coordinates detected!
re-calculating split coordinates for these chromosomes
Use:
/toy_output/RECAL1_FINAL_SPLIT_coord_for_GTF.bed as the input to modify GTF file.
and:
/toy_output/RECAL1_FINAL_SPLIT_coord_for_FASTA.bed
 as the input for splitting FASTA.
```


##### 2) Validate_inputs 
```
centromere_path2="/toy_input/IN0_toy_centromeres_for_gtf.bed" ### Depending on the input data, the incput might be /toy_input/RECAL1_FINAL_SPLIT_coord_for_GTF.bed
validate_inputs(
    gtf_path, 
    centromere_path2)
```

#### 3) Process GTF
```
centromere_path2="/toy_input/IN0_toy_centromeres_for_gtf.bed"
unchar_region="CAJHDD"
mito="Mt"
chloro="Pt"
processed_GTF_First_Fith_Sixth_Col = process_gtf(
    gtf_path, 
    centromere_path2, 
    unchar_region, 
    mito, 
    chloro
    )
```

#### 4) Process 9th column of the GTF file
```
keep_attributes= c("gene_id", "transcript_id", "gene_name")
out_path="/toy_output/"
genome_name ="BarleyMOREXV3_toy"
genome_version ="v51_ENSEMBL"
final_result = process_attributes_column(
    processed_GTF_First_Fith_Sixth_Col, 
    keep_attributes,
    out_path,
    genome_name,
    genome_version
    )
```  







## 5. Session Info  

```{r session-info}
sessionInfo()
```



## 6. Acknowledgements  

This work was supported by resources provided by the Pawsey Supercomputing Research Centres Nimbus Research Cloud (https://doi.org/10.48569/v0j3-qd51), with funding from the Australian Government and the Government of Western Australia.