---
title: "scShardSplitRef"
author: "Irina Kuznetsova"
output: rmarkdown::html_vignette    ####html_vignette
vignette: >
  %\VignetteIndexEntry{scShardSplitRef}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  )
devtools::load_all() #### delete after publishing
```

## 1. Purpose

This synthetic dataset provides a minimal example to demonstrate the logic of the algorithm that prepares reference files (GTF and FASTA) for `cellranger‑arc mkref`. It allows users to test the workflow without needing large genome files and helps confirm that the proposed workflow is functioning correctly. It also provides an example of how to use it.

## 2. {scShardSplitRef} guided walk through using a synthetic example

### Installation

```{r setup}
library("scShardSplitRef")
library("knitr") # used for a kable table later in the document
```

### Prerequisites

| Software | Version | Purpose             |
|----------|---------|---------------------|
| bedtools | ≥ 2.30  | BED file operations |

### Input files

| File | Format | Description | File Name |
|------------------|------------------|------------------|------------------|
| Gene annotations | GTF | Gene features and coordinates | extdata/AlgorithmSynthetic.gtf |
| Centromere coordinates | BED-like | Centromere positions per chromosome (x) | extdata/AlgorithmSynthetic.bed |




## 3. Algorithm explonatation

The first function called `validate_inputs()`, checks the correct format of (GTF and BED-like) input files.

-   **BED:** 3 columns, tab-delimited, no header, no comments format (this centromere coordinates file is created manually)
-   **GTF**: 9 columns, tab-delimited, lines beginning with `#` at the top of the file are treated as comments and allowed
-   All files need to be decompressed

#### The help page for the validate_inputs()

```{r, eval=FALSE}
?validate_inputs
```

#### Validate (FASTA and BED) input files

```{r}
bed_path <- system.file(
  "extdata/AlgorithmSynthetic.bed",
  package = "scShardSplitRef" )

gtf_path <- system.file(
  "extdata/AlgorithmSynthetic.gtf",
  package = "scShardSplitRef"  )

# Validate input (FASTA and BED) files
input_verif <- validate_inputs(bed = bed_path, 
                               gtf = gtf_path )
kable(input_verif$bed)
kable(input_verif$gtf[1:5,])
```

> If the GTF and BED formats are correct, a confirmation message will be printed to the console: *"All files loaded and validated successfully"*



The next function called ```determine_split_regions()``` prepares split coordinates in appropriate BED-like format. This function ensures that chromosomes are not split in the middle of a gene region.

### The help page for the determine_split_regions() function

```{r, eval=FALSE}
?determine_split_regions
```


**Input data** 
This function takes a BED file and a GTF file as input.


**Parameters:**

- `limit`: can be left as the default, which is the Cell Ranger ARC limit 
of 2²⁹ (536,870,912 bp). For the purpose of this tutorial, to better 
understand the algorithm, we set `limit=500`
- `shift_by`: the step size used to move the split coordinate to the right 
when searching for a safe location outside a genic region
- `clearance`: a buffer that checks whether there is a clear region free of 
genes on both sides of the proposed split coordinate

In this example we use:
```
limit = 500 
shift_by = 30 
clearance = 40
```

```{r}
# Define input files directory
bed_path <- system.file("extdata/AlgorithmSynthetic.bed",
  package = "scShardSplitRef" )

gtf_path <- system.file("extdata/AlgorithmSynthetic.gtf",
  package = "scShardSplitRef"  )

# Define output directory
output_file <- file.path(tempdir(), 
                         "outdata")
dir.create(output_file, 
           recursive = TRUE, 
           showWarnings = FALSE           )
output_file_name_dir <- file.path(output_file, 
                                  "DetermineSplitReg_AlgorithmSynthetic.gtf")

# Determine split regions
get_split_reg <- determine_split_regions(
  bed = bed_path,
  gtf = gtf_path,
  output_bed = output_file_name_dir,
  limit = 500,
  shift_by = 30L,
  clearance = 40
  )

get_split_reg

```


**How it works:** 
The function checks whether the centromere location from the BED file overlaps with a genic region.  
**No overlap:** the chromosome is split at that coordinate, producing two regions:

- Region 1: from the beginning of the chromosome to the split coordinate
- Region 2: from the split coordinate to the end of the chromosome

<details>
<summary><b>Visual (chr1): the chromosome is split at centromere coordinate</b></summary>
```
chr1:	0-1000
limit = 500 
shift_by = 30 
clearance = 40

gene[G1]:	100-200
```

      chr1: 0    100  200       500            1000
            |-----|====|---------|--------------|
                  [G1]           ^
                                 |
                              split (500)
      
            |<---- Region 1 ---->|<-- Region 2 ->|
            
            Final: split at 500 -> 1 split cooridnate (2 shards)   

</details>


**Overlap detected:** the coordinate needs to be moved to a safe location that does not overlap any genes. The algorithm proceeds as follows:

The split position is calculated from the limit parameter. It looks to see if there is sufficient clearance between the split point and any neighbouring genes left or right. If the clearance test does not pass the function uses ```shift_by```, which essentially defines the size of the steps to iteratively take as it tries to walk away from any genes that are too close.  

> Note: ```clearance``` and ```shift_by``` operate independently. ```clearance``` determines whether a position is acceptable; ```shift_by``` determines how far to move when it is not.

<details>

<summary><b>Visual (chr2): the initial split coordinate is inside of a gene</b></summary>

```
chr2:	0-1000
limit = 500 
shift_by = 30 
clearance = 40

gene[G2]: 450-550
```

      chr2: 0              450      550            1000
            |----------------|=======|--------------|
                             [G2] ^
                                  |
                               split (500)
                               
            Initial split coordinate 500 -> falls inside G2 [450–550]
            Shift right by 30 bp -> 530 -> still inside G2 [450–550]
            Shift right by 30 bp -> 560 -> outside G2 [450–550], NOT passing clearance = 40
            Shift right by 30 bp -> 590 -> outside G2 [450–550], PASSING clearance = 40, BUT exceeds limit = 500
              Subdivide: new split coordinate = 590 / 2 = 295
              Check clearance = 40 around proposed coordinate 295 → PASSING
              Split coordinate = 295
      

      chr2: 0               450      550            1000
            |----------------|=======|--------------|
                    ^           [G2]      
                    |                    
                split (295)                
                
            
            As the distance between 295 and 1000 exceeds the limit of 500, find the next split coordinate:
            Next split coordinate 295 + 295 = 590 — outside G2 [450–550], PASSING clearance = 40
            Split coordinate = 590
                                               

      chr2: 0              450      550            1000
            |----------------|=======|--------------|
                    ^           [G2]     ^
                    |                    |
                split (295)             split (590)
                
            Final: split at 295 and 590 -> 2 split cooridnates (3 shards)    
                

</details>





<details>

<summary><b>Visual (chr3): two overlapping genes</b></summary>
```
chr3:	0-1000
limit = 500 
shift_by = 30 
clearance = 40

gene[G3L]:	460-510
gene[G3R]:	490-560
```

      chr3: 0              460  490     510   560             1000
            |----------------|===|overlap|======|--------------|
                             [G3L]   ^    [G3R]
                                     |
                               split (500)
                               
            Initial split coordinate 500 is inside of the G3L[460-510] and G3R[490-560]
            Shift to the right by 30bp -> 530 - inside of the G3R[490-560]
            Shift to the right by 30bp -> 550 - still inside of the G3R[490-560]
            Shift to the right by 30bp -> 590 - outside of the G3R[490-560], not passing clearance
            Shift to the right by 30bp -> 620 - outside of the G3R[490-560] -> PASSING clearance
            Split coordinate = 620
           
            
      chr3: 0              460  490     510   560             1000
            |----------------|===|overlap|======|--------------|
                             [G3L]        [G3R]       ^ 
                                                      |
                                                split (620)
            
            As the distance between 0 and 620 exceeds the limit of 500, find the next split coordinate: 
            Next split coordinate 620/2 = 310 —> outside G3L[460-510] and G3R[490-560], PASSING clearance = 40
            Split coordinate = 310
            
            
      chr3: 0              460  490     510   560             1000
            |----------------|===|overlap|======|--------------|
                      ^      [G3L]        [G3R]       ^ 
                      |                               |
                 split (310)                       split (620)    
            
            Final: split at 620 and 310 -> 2 split cooridnates (3 shards)

</details>





<details>

<summary><b> Visual (chr4): one gene in each arm (no collision)</b></summary>

```
chr4:	300-1500 
Note, that each chromosome is defined by a coordinate range starting at 0. If a feature's start position is not 0, it indicates a split coordinate. In this example, 300 is the split coordinate, dividing the chromosome into two segments: 0–300 and 300–500.
limit = 500 
shift_by = 30 
clearance = 40

gene[G4L]:	50	100
gene[G4R]:	800	900
```

      chr4: 0     50  100                        800   900     1500
            |-----|===|--------------------------|======|------|
                  [G4L]         ^                 [G4R]
                                |     
                               split (300)
                               
            Initial split coordinate 300 -> fulfils limit setting 
            As the distance between 300 and 1500 exceeds the limit of 500, find how many shards: 
            1200/500 ~ (3 shards) -> 1200/4=400
              Next split coordinate 300+400=700, outside of G4R[800-900], PASSING clearance = 40
              Next split coordinate 700+400=1100, outside of G4R[800-900], PASSING clearance = 40

      chr4: 0     50  100                        800   900     1500
            |-----|===|--------------------------|======|------|
                  [G4L]         ^             ^   [G4R]     ^
                                |             |             |
                            split (300)    split (700)    split (1100)
           
      Final: split at 300, 700 and 1100 -> 3 split cooridnates (4 shards)

</details>


<details>

<summary><b> Visual (chr5): a split coordinate cuts through a gene</b></summary>
```
chr5:	600-1500
# Note, that each chromosome is defined by a coordinate range starting at 0. If a feature's start position is not 0, it indicates a split coordinate. In this example, 600 is the split coordinate, dividing the chromosome into two segments: 0–600 and 600–1500.

limit = 500 
shift_by = 30 
clearance = 40

gene[G5BS]:	550-650
```

      chr5: 0                550    650                               1500
            |-----------------|======|---------------------------------|
                                 ^ [G5BS]                
                                 |     
                               split (600)
                               
            
            As the distance between 0 and 600 exceeds the limit of 500, find how many shards: 600/500 = 2
            600/2 = 300 
            As the distance between 600 and 1500 exceeds the limit of 500, find next split cooridnate
            (1500-600)/2 = 600 + 450 = 1050 
            
            
      chr5: 0                550    650                               1500
            |-----------------|======|---------------------------------|
                     ^           ^  [G5BS]        ^     
                     |           |       |        |
                 split (300)   split (600)    split (1050)
            
            
            Initial split coordinate 600 -> is inside of the G5BSL[550-650] and exceeds allowed limit 500
            Shift to the right by 30bp -> 630 - inside of the G5BSL[550-650]
            Shift to the right by 30bp -> 660 - outside of the G5BSL[550-650], not passing clearance
            Shift to the right by 30bp -> 690 - outside of the G3R[490-560] -> PASSING clearance
            Split coordinate = 690
            
            
     chr5: 0                550    650                               1500
            |-----------------|======|---------------------------------|
                     ^         [G5BS]     ^        ^
                     |                    |        |
                 split (300)          plit (690)  split (1050)
            
            Final: split at 300, 690 and 1050 -> 3 split cooridnates (4 shards)
</details>



<details>

<summary><b> Visual (chr6): gene structure with exons, CDS, and transcript: gene overlaps with cutting coordinate</b></summary>
```
chr6:	0-1000
limit = 500 
shift_by = 30 
clearance = 40

gene[G6R]:        400-500
transcript[G6R]:  400-500
exon[G6R]:        400-450
exon[G6R]:	      460-480
exon[G6R]:        490–500
CDS[G6R]:	        410-495
five_prime_utr[G6R]:  400-409

      chr6: 0                                400                    500                      1000
            |--------------------------------|======================|------------------------|
                                                    [G5R]           ^
                                                                    |
                                                                split (500)       
            
            The chromosome 6 has no proposed centromer coordinate (reasone: for e.g. not accessible from databases yet) 
              The allowed limit is 500bp
              
            
      chr6: 0                                400                    500                      1000
            |--------------------------------|======================|------------------------|
                                                    [G5R]           ^
                                                                    |
                                                                split (500)    
                                                                
            
            Split coordinate 500 -> is on end cooridnate of the G5R[100-550]:
              Shift to the right by 30bp -> 530 - outside of the G5R[100-550], not passing clearance
              Shift to the right by 30bp -> 560 - outside of the G5R[100-550] -> PASSING clearance
            
      chr6: 0                                400                    500                      1000
            |--------------------------------|======================|------------------------|
                                                    [G5R]               ^
                                                                        |
                                                                     split (560)    
                                                                     
            
            As the distance between 0 and 560 exceeds the limit of 500, find how many shards: 560/500 ~ 2
            560/2 = 280
            
            
      chr6: 0                                400                    500                      1000
            |--------------------------------|======================|------------------------|
                              ^                     [G5R]               ^
                              |                                         |
                           split (280)                              split (560) 
                                                                    
            Final: split at 280, 560 -> 2 split cooridnates (3 shards)   
                                                                    
                                                       

```
</details>






## 3. Session Info

```{r session-info}
sessionInfo()
```

## 4. Acknowledgements

This work was supported by resources provided by the Pawsey Supercomputing Research Centres Nimbus Research Cloud (<https://doi.org/10.48569/v0j3-qd51>), with funding from the Australian Government and the Government of Western Australia.
