---
title: "scShardSplitRef"
author: "Irina Kuznetsova"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scShardSplitRef}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
devtools::load_all()  #### delete after publishing
```

## Package overview

*scShardSplitRef* designed for researchers working with multiome single-cell data from species that require building a custom reference genome, particularly when one or more chromosomes/contigs in the reference FASTA exceed the Cell Ranger ARC limit of 536.8Mb (2\^29 bases). *scShardSplitRef* enables splitting of long chromosomes/contigs generating shorter fragments that are compitable with Cell Ranger ARC requirements. The software outputs FASTA and GTF files containing the split chromosomes/contigs, which can be directly processed by `cellranger-arc mkref`.

To familiarise yourself with the package and workflow of preparing reference genome sequence (FASTA file) and gene annotations (GTF file) for Cell Ranger ARC we suggest starting with a guided walk through. We are using barley genome (MOREX V3) as an example genome. It starts with GTF file and BED are available from our GitHub page from `inst/extdata`.

## Setup the scShardSplitRef package

For this tutorial, we will be using a genome sequence (FASTA file) and gene annotations (GTF file) of the barley genome (*Hordeum Vulgare, Morex V3*) available from [Ensembl](https://ftp.ensemblgenomes.ebi.ac.uk) and centromere coordinates from [GrainGenes](https://graingenes.org/GG3/content/morex-v3-files-2021).

We begin by assessing the formatting of the input GTF and BED files provided by the user, followed by checking and, if necessary, modifying the split coordinates in the BED file. The GTF file is then updated according to the split position. Finally, FASTA file is split and formatted for compatibility with Cell Ranger ARC.

```{r}
# Install the package
#install.packages("pak")

# Load the package 
library("dplyr")
library("scShardSplitRef")

# Load all the code in an R package you're currently developing
#devtools::load_all() 

# Download GTF, FASTA and centromere coordinates files
# Note1: all files need to be decompressed
# Note2: use correctly formatted BED file with centromere coordinates 
```

<details>

<summary><b>**What do GTF and centromere files for MOREX V3 look like?**</b></summary>

All files need to be decompressed.

```{r}
### Gene annotations (GTF) file: "/inst/extdata/01_Hordeum_vulgare.Mt_Pt_v49_Barley_v62_COMBINED.gtf"
gtf_file <- system.file("extdata", "01_Hordeum_vulgare.Mt_Pt_v49_Barley_v62_COMBINED.gtf", package = "scShardSplitRef")
gtf_df <- read.table(file = gtf_file,
                     header = FALSE, 
                     sep = "\t")
dim(gtf_df)     ### 524562 x 9
gtf_df[1:10, ]
```

The GTF file contains 9 columns.

```         
#!genome-build MorexV3_pseudomolecules_assembly
#!genome-version MorexV3_pseudomolecules_assembly
#!genome-date 2021-04
#!genome-build-accession GCA_904849725.1
#!genebuild-last-updated 2021-04
2H      IPK     gene    601169528       601171444       .       -       .       gene_id "HORVU.MOREX.r3.2HG0191020"; gene_source "IPK"; gene_biotype "protein_coding";
2H      IPK     transcript      601169528       601171444       .       -       .       gene_id "HORVU.MOREX.r3.2HG0191020"; transcript_id "HORVU.MOREX.r3.2HG0191020.1"; gene_source "IPK"; gene_biotype "protein_coding"; transcript_source "IPK"; transcript_biotype "protein_coding"; tag "Ensembl_canonical";
2H      IPK     exon    601169754       601171444       .       -       .       gene_id "HORVU.MOREX.r3.2HG0191020"; transcript_id "HORVU.MOREX.r3.2HG0191020.1"; exon_number "1"; gene_source "IPK"; gene_biotype "protein_coding"; transcript_source "IPK"; transcript_biotype "protein_coding"; exon_id "HORVU.MOREX.r3.2HG0191020.1-E1"; tag "Ensembl_canonical";
2H      IPK     CDS     601169754       601171444       .       -       0       gene_id "HORVU.MOREX.r3.2HG0191020"; transcript_id "HORVU.MOREX.r3.2HG0191020.1"; exon_number "1"; gene_source "IPK"; gene_biotype "protein_coding"; transcript_source "IPK"; transcript_biotype "protein_coding"; protein_id "HORVU.MOREX.r3.2HG0191020.1"; tag "Ensembl_canonical";
2H      IPK     start_codon     601171442       601171444       .       -       0       gene_id "HORVU.MOREX.r3.2HG0191020"; transcript_id "HORVU.MOREX.r3.2HG0191020.1"; exon_number "1"; gene_source "IPK"; gene_biotype "protein_coding"; transcript_source "IPK"; transcript_biotype "protein_coding"; tag "Ensembl_canonical";
```

```{r}
### Centromere file: "/inst/extdata/02_MorexV3_centromere_coordinates.bed"
bed_file <- system.file("extdata","02_MorexV3_centromere_coordinates.bed" , package = "scShardSplitRef")
centr_bed_df <- read.table(file = bed_file, 
                           header = FALSE, 
                           sep = "\t")
dim(centr_bed_df)  ## 292 x 3
centr_bed_df[1:10, ]
```

The format of the centromere coordinates file is a tab-delimited 3-column file, where the 1st column contains chromosome/contig names, the 2nd column contains the centromere coordinate, and the 3rd column contains chromosome/contig region end coordinates.

```         
1H      206486643       516505932
2H      301293086       665585731
3H      267852507       621516506
4H      276149121       610333535
5H      204878572       588218686
6H      256319444       561794515
7H      328847192       632540561
CAJHDD010000004.1       0       972986
CAJHDD010000001.1       0       680018
CAJHDD010000005.1       0       517010
```

</details>

## scShardSplitRef processing workflow

##### Step-1

The initial step is to check if required files are provided in the correct format.

```{r}
bed_path <- system.file("extdata", "02_MorexV3_centromere_coordinates.bed", package = "scShardSplitRef")
gtf_path <- system.file("extdata", "01_Hordeum_vulgare.Mt_Pt_v49_Barley_v62_COMBINED.gtf", package = "scShardSplitRef")
input_verif <- validate_inputs(bed = bed_path, 
                               gtf = gtf_path)
```

If GTF and BED formats are correct, a confirmation message will be printed to the console.

```         
All files loaded and validated successfully.
```

##### Step-2

The next step is to prepare split coordinates in BED format, ensuring that chromosomes are not split in the middle of a gene region. To do this, we first test whether our centromere coordinates overlap with any gene features.

```{r}
# Build BED-like split intervals from genomic regions (BED) and gene annotations (GTF)
bed_path <- system.file("extdata", "02_MorexV3_centromere_coordinates.bed", package = "scShardSplitRef")
gtf_path <- system.file("extdata", "01_Hordeum_vulgare.Mt_Pt_v49_Barley_v62_COMBINED.gtf", package = "scShardSplitRef")
out_file <- system.file("outdata", "B1_Split_regions.bed", package = "scShardSplitRef")
get_split_reg <- determine_split_regions(bed = bed_path, 
                        gtf = gtf_path, 
                        output_bed = out_file, 
                        limit = 2L^29L, 
                        shift_by = 20L)
```

<details>

<summary><b>**What do the results "B1_Split_regions.bed" looks like?**</b></summary>

```         
1H  0   206486643
1H  206486643   516505932
2H  0   301293086
2H  301293086   665585731
3H  0   267852507
3H  267852507   621516506
4H  0   276149121
4H  276149121   610333535
5H  0   204878572
5H  204878572   588218686
6H  0   256319444
6H  256319444   561794515
7H  0   328847192
7H  328847192   632540561
CAJHDD010000004.1   0   972986
CAJHDD010000001.1   0   680018
```

</details>

##### Step-3

Generate gene annotation (GTF) file in compatible to Cell Ranger ARC format, wher the first column of GTF file should be in the following form `Chr-RegionStart-RegionEnd`. None that the output file name is hardcoded and starts with *"B1_FINAL_MODIFIED_GTF..."*

```{r}
### GTF_FIN_OUT="/outdata/"
split_regions_bed_path <- system.file("outdata", "B1_Split_regions.bed", package = "scShardSplitRef")
gtf_path <- system.file("extdata", "01_Hordeum_vulgare.Mt_Pt_v49_Barley_v62_COMBINED.gtf", package = "scShardSplitRef")
out_file2 <- system.file("outdata", package = "scShardSplitRef")
GTF_processed <- process_gtf(split_regions_bed  = "/inst/outdata/B1_Split_regions.bed", 
                          gtf = gtf_path, 
                          genome_name = "barley", 
                          genome_version = "3", 
                          out_path = out_file2)

```

<details>

<summary><b>**What do the results in "B1_FINAL_MODIFIED_GTF_barley_3.gtf" looks like?**</b></summary>

```         
2H-301293086-665585731  IPK     gene    299876443       299878359       .       -       .       gene_id "HORVU.MOREX.r3.2HG0191020"; gene_source "IPK"; gene_biotype "protein_coding";
2H-301293086-665585731  IPK     transcript      299876443       299878359       .       -       .       gene_id "HORVU.MOREX.r3.2HG0191020"; transcript_id "HORVU.MOREX.r3.2HG0191020.1"; gene_source "IPK"; gene_biotype "protein_coding"; transcript_source "IPK"; transcript_biotype "protein_coding"; tag "Ensembl_canonical";
2H-301293086-665585731  IPK     exon    299876669       299878359       .       -       .       gene_id "HORVU.MOREX.r3.2HG0191020"; transcript_id "HORVU.MOREX.r3.2HG0191020.1"; exon_number "1"; gene_source "IPK"; gene_biotype "protein_coding"; transcript_source "IPK"; transcript_biotype "protein_coding"; exon_id "HORVU.MOREX.r3.2HG0191020.1-E1"; tag "Ensembl_canonical";
2H-301293086-665585731  IPK     CDS     299876669       299878359       .       -       0       gene_id "HORVU.MOREX.r3.2HG0191020"; transcript_id "HORVU.MOREX.r3.2HG0191020.1"; exon_number "1"; gene_source "IPK"; gene_biotype "protein_coding"; transcript_source "IPK"; transcript_biotype "protein_coding"; protein_id "HORVU.MOREX.r3.2HG0191020.1"; tag "Ensembl_canonical";
```

</details>

##### Step-4

Final steps is tor split FASTA file and formatted to be compitable with Cell Ranger ARC

```{linux}
#Please note that due to file size limitations, the barley genome FASTA file is not available on our GitHub repository at /inst/extdata/
IN_FA = "/inst/extdata/01_Hordeum_vulgare.MorexV3_pseudomolecules_assembly.dna.primary_assembly_COMBINED_Mt_Pt_v49_Barley_v62.fa"
BED_FIN_OUT  = "/inst/outdata/B1_Split_regions.bed"
FASTA_FIN_OUT = "/inst/outdata/"

# Split FASTA 
bedtools getfasta -fi ${IN_FA} -bed ${BED_FIN_OUT} -fo ${FASTA_FIN_OUT}/B1_FASTA_split.fasta

# Convert FASTA heading in Cell Ranger ARC compatible format: Chr-RegionStart-RegionEnd
sed '/^>/s/:/-/g' ${FASTA_FIN_OUT}/B1_FASTA_split.fasta > ${FASTA_FIN_OUT}/B1_FASTA_split_final.fasta

```

<details>

<summary><b>**What do the results in "B1_FASTA_split_final.fasta" looks like?**</b></summary>

```         
>1H-0-206486643
ATTTGTAGTGCTTTTCAATTTCAGGGTCAA.....
>1H-206486643-516505932
ACTGGCCAAAATAGATCAAAATTGCGAGTTTTGACGAGTTCCCCGTAAGCGGACTTCGG...........
>2H-0-301293086
GTGAGAATAGGCC.........
>2H-301293086-665585731
```

</details>

##### Step-5

Cell Ranger ARC

#### prepaper confif file

```         
{
    organism: "barley"
    genome: ["MOREX3_v3"]
    input_fasta: ["B1_FASTA_split_final.fasta"]
    input_gtf: ["/outdata/B1_FINAL_MODIFIED_GTF_barley_3.gtf"]
    non_nuclear_contigs: ["Mt-0-525599", "Pt-0-115974"]

}
```

#### run Cell Ranger

```         
cellranger-arc mkref --config=${CONFIG_IN} --nthreads 24 > mkref.log 2>&1
```

## Acknowledgements

This work was supported by resources provided by the [Pawsey Supercomputing Research Centre's](https://pawsey.org.au/) Nimbus Research Cloud (<https://doi.org/10.48569/v0j3-qd51>), with funding from the Australian Government and the Government of Western Australia.

GRDC AAGI project support code: AAGI–CU–RD–SNB #TODO: Is there any standardised GRD message for acknowledgment?
