---
title: "scShardSplitRef: Getting Started"
author: "Irina Kuznetsova"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scShardSplitRef: Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Package Overview

{scShardSplitRef} designed for researchers working with multiome single-cell data from species that require building a custom reference genome, particularly when one or more chromosomes/contigs in the reference FASTA exceed the Cell Ranger ARC limit of `536.8Mb (2\^29 bases)`.
{scShardSplitRef} enables splitting of long chromosomes/contigs generating shorter fragments that are compatible with Cell Ranger ARC requirements.
The software outputs FASTA and GTF files containing the split chromosomes/contigs, which can be directly processed by `cellranger-arc mkref`.

To familiarise yourself with the package and workflow of preparing reference genome sequence (FASTA file) and gene annotations (GTF file) for Cell Ranger ARC we suggest starting with the [Guided Walk-Through](#guided-walk-through).
In this tutorial we are using barley genome (MOREX V3).
Example data files are distributed with the package for example purposes.

## Guided Walk-Through {#guided-walk-through}

### Installation

```{r setup, eval=TRUE, echo=FALSE, include=FALSE}
library("knitr") # used for a kable table later in the document
```

### Prerequisites

| Software        | Version | Purpose                         |
|-----------------|---------|---------------------------------|
| bedtools        | ≥ 2.30  | BED file operations             |
| Cell Ranger ARC | ≥ 2.0   | Single-cell multiome processing |

### Input Files

| File | Format | Description | Comment |
|----------------|----------------|---------------------|---------------------|
| Gene annotations | GTF | Gene features and coordinates | Morex v3, subset to 2 chromosomes |
| Centromere coordinates | BED-like | Centromere positions per chromosome (x1) | 2 chromosomes |
| Genome sequence | FASTA | Reference genome sequence | (x2) Not bundled; used for Cell Ranger ARC demo only; download from Ensembl |

> **(x1) Note1:** If centromere coordinates are not available for your species, the second column of the BED-like file should be set to `0`.
> If centromere coordinates are available, the second column should be set to the centromere coordinate.
> {scShardSplitRef} will then check for any overlaps with genic regions and generate the correct GTF file that is compitable with `cellranger-arc mkref`.
>
> **(x2) Note2:** Please note that the FASTA file is not required for the package, but is used to demonstrate FASTA preparation for the Cell Ranger ARC command in the workflow description.
> It is not included in the repository due to the large file size; download from Ensembl.

<br>

## Step-by-Step Guide

For this tutorial, we will be using a subset of barley genome (*Hordeum Vulgare, Morex V3*).
We begin by downloading the FASTA and GTF files from [Ensembl](https://ftp.ensemblgenomes..ebi.ac.uk) and creating a tab-delimited BED-like file with centromere coordinates from [GrainGenes](https://graingenes.org/GG3/content/morex-v3-files-2021).
We then assess the formatting of the input GTF and BED files, followed by checking and, if necessary, modifying the split coordinates in the BED file.
The GTF file is subsequently updated according to the split position.
Finally, the FASTA file is split and formatted for compatibility with Cell Ranger ARC. All files must be decompressed prior to use.

<br>

<details>

<summary><b>What does a GTF file for MOREX V3 look like?`<`{=html}
`/b></summary>`{=html}

```{r, eval=TRUE}
# Gene annotations (GTF) file. The format of the GTF file is a tab-delimited 9-column file.
# data.dir = "/inst/extdata/morex3_subset_1H_2H.gtf.gz"

gtf_file <- system.file(
  "extdata/morex3_subset_1H_2H.gtf.gz",
  package = "scShardSplitRef"
)
decompress_gtf <- gzfile(gtf_file, open = "rt")
gtf_df <- read.table(file = decompress_gtf, header = FALSE, sep = "\t")
close(decompress_gtf)
dim(gtf_df)
gtf_df[1:10, ]
```

</details>

<details>

<summary><b>What does a BED-like centromere file for MOREX V3 look like?`<`{=html}
`/b></summary>`{=html}

```{r, eval=TRUE}
# The format of the centromere coordinates file is a tab-delimited 3-column file, where
#   the 1st column contains chromosome/contig names,
#   the 2nd column contains the centromere coordinate (or 0 if this information is not available for your species),
#   the 3rd column contains the chromosome/contig end coordinates.
# Manually formatted centromere file (please check the expected format below):
# data.dir = "/inst/extdata/morex3_subset_1H_2H.bed"

bed_file <- system.file(
  "extdata/morex3_subset_1H_2H.bed",
  package = "scShardSplitRef"
)
centr_bed_df <- read.table(file = bed_file, header = FALSE, sep = "\t")
dim(centr_bed_df)
centr_bed_df
```

</details>

### Step-1

The goal of the `determine_split_regions()` function is to create appropriate split regions.
It initially checks FASTA and BED-like input files, then prepares split coordinates in the tab-delimited BED-like format.
This function ensures that chromosomes are not split in the middle of a gene region.
The BED-like file with centromere coordinates if available, or zeros if not.
This file is manually created.
All files need to be decompressed.

### The Help Page for the `determine_split_regions()` Function

``` r
?determine_split_regions
```

### Check formatting of input files and build BED-like split intervals

```{r, eval=TRUE}
# Load the library
library(scShardSplitRef)

# Load MOREX V3 (subset to 1H, 2H)
# Note1: all files need to be decompressed for use in scShardSplitRef and Cell Ranger ARC
# Note2: use correctly formatted BED file with centromere coordinates
# Task: Check formatting of input files. Build BED-like split intervals using genomic regions (BED) and gene annotations (GTF) input files

# Read GTF file
gtf_path <- system.file(
  "extdata/morex3_subset_1H_2H.gtf.gz",
  package = "scShardSplitRef"
)
tmp_gtf_path <- tempfile(fileext = ".gtf")
con_in <- gzcon(file(gtf_path, open = "rb"))
con_out <- file(tmp_gtf_path, open = "wb")
writeBin(readBin(con_in, raw(), n = 1e8), con_out)
close(con_in)
close(con_out)

# Read BED file
bed_path <- system.file(
  "extdata/morex3_subset_1H_2H.bed",
  package = "scShardSplitRef"
)

# Define output path
output_file <- file.path(tempdir(), "split")
dir.create(output_file, recursive = TRUE, showWarnings = FALSE)
output_file_name_dir <- file.path(
  output_file,
  "morex3_subset_1H_2H_Split_Regions.bed"
)

# Determine split regions
get_split_reg <- determine_split_regions(
  bed = bed_path,
  gtf = tmp_gtf_path,
  output_bed = output_file_name_dir,
  limit = 2L^29L,
  shift_by = 20L,
  clearance = 20L
)

get_split_reg
```

<details>

<summary><b>What does the generated BED file look like?`<`{=html}
`/b></summary>`{=html}

```{r, eval=TRUE}
bed_f <- system.file(
  "extdata/morex3_subset_1H_2H_Split_Regions.bed",
  package = "scShardSplitRef"
)
split_region_bed_df <- read.table(file = bed_f, header = FALSE, sep = "\t")
dim(split_region_bed_df)

split_region_bed_df
```

</details>

### Step-2

Generate a gene annotation (GTF) file in compatible to Cell Ranger ARC format.
The first column of the GTF file should be in the following form `Chr-RegionStart-RegionEnd`.
Note that the output file name is hardcoded and starts with "B1_FINAL_MODIFIED_GTF\_[*genome_name*]\_[*genome_version*].gtf".

#### The Help Page for the `process_gtf()` Function

``` r
?process_gtf
```

#### Generate Gene Annotation (GTF) File in Compatible to Cell Ranger ARC Format

```{r, eval=TRUE}
# Process GTF file and write file to tempdir()
print(output_file_name_dir)
print(tmp_gtf_path)
process_gtf(
  split_regions_bed = output_file_name_dir,
  gtf = tmp_gtf_path,
  genome_name = "subset_barley",
  genome_version = "3",
  out_path = tempdir()
)
```

> The output file is in the GTF format.
> Note that the first column has been reformatted, the regions are now split.\
> The chromosome name is formatted as `1H-0-206486643`, which is the region from the beginning of the chromosome to the split position;\
> `1H-206486643-516505932` is the region from the split position to the end of the chromosome.

<details>

<summary><b>What does the generated GTF file look like?`<`{=html}
`/b></summary>`{=html}

```{r, eval=TRUE}
gtf_modified <- list.files(
  tempdir(),
  pattern = "B1_FINAL_MODIFIED_GTF_subset.*\\.gtf$",
  full.names = TRUE
)

processed_gtf <- read.table(file = gtf_modified, header = FALSE, sep = "\t")
print(gtf_modified)
dim(processed_gtf)

sep <- as.data.frame(matrix("...", nrow = 1, ncol = 5))
colnames(sep) <- colnames(processed_gtf)[1:5]

kable(rbind(
  processed_gtf[200:203, 1:5],
  sep,
  processed_gtf[(nrow(processed_gtf) - 2):nrow(processed_gtf), 1:5]
))
```

</details>

<br> **{scShardSplitRef} generated a GTF file with split coordinates that can be used in Cell Ranger ARC.** <br>

### Step-3

Here we show the workflow on obtaining FASTA files from the Ensembl FTP server, explaining how to split FASTA and keep the header naming convention in the same style as the GTF file generated by the `scShardSplitRef` package.

``` bash
# Get FASTA from Ensembl
# Hordeum vulgare, (release-62), Ensembl
wget https://ftp.ensemblgenomes.ebi.ac.uk/pub/plants/release-62/fasta/hordeum_vulgare/dna/Hordeum_vulgare.MorexV3_pseudomolecules_assembly.dna.primary_assembly.1H.fa.gz
wget https://ftp.ensemblgenomes.ebi.ac.uk/pub/plants/release-62/fasta/hordeum_vulgare/dna/Hordeum_vulgare.MorexV3_pseudomolecules_assembly.dna.primary_assembly.2H.fa.gz
```

``` bash
# Decompress FASTA
gunzip Hordeum_vulgare.MorexV3_pseudomolecules_assembly.dna.primary_assembly.1H.fa.gz
gunzip Hordeum_vulgare.MorexV3_pseudomolecules_assembly.dna.primary_assembly.2H.fa.gz
```

``` bash
# Concatenate into one file
cat Hordeum_vulgare.MorexV3_pseudomolecules_assembly.dna.primary_assembly.1H.fa Hordeum_vulgare.MorexV3_pseudomolecules_assembly.dna.primary_assembly.2H.fa >morex3_subset_1H_2H.fasta
```

``` bash
# Please note that due to the file size limitations, the barley genome FASTA file is not available on our GitHub repository at /inst/extdata/
IN_FA="/data/morex3_subset_1H_2H.fasta"
BED_FIN_OUT="/inst/extdata/morex3_subset_1H_2H_Split_Regions.bed"
FASTA_FIN_OUT="/data/"

# Split FASTA
bedtools getfasta -fi ${IN_FA} -bed ${BED_FIN_OUT} -fo ${FASTA_FIN_OUT}/morex3_SPLIT_1H_2H.fasta

# Note, FASTA heading should be in `Chr-RegionStart-RegionEnd` format to be compitable with Cell Ranger ARC
sed '/^>/s/:/-/g' ${FASTA_FIN_OUT}/morex3_SPLIT_1H_2H.fasta >${FASTA_FIN_OUT}/morex3_split_FINAL_1H_2H.fasta

# Validate
seqkit fx2tab --length --name ${FASTA_FIN_OUT}/morex3_split_FINAL_1H_2H.fasta

1H-0-206486643 206486643
1H-206486643-516505932 310019289
2H-0-301293086 301293086
2H-301293086-665585731 364292645
```

<details>

<summary><b>What does the generated FASTA file look like?`<`{=html}
`/b></summary>`{=html}

```         
# morex3_split_FINAL_1H_2H.fasta

>1H-0-206486643
GCATTTGGTTGCTTTTATTCTAAACTA...
>1H-206486643-516505932
ATCAGTACATGAGCTAGTATTAGAGGCTAGACTAGGGATCAATTGGTGTTTGTA...
>2H-0-301293086
GCCCTGAC...
```

</details>

### Step-4

The final step demonstrates all required input files for the Cell Ranger ARC command.
The genome sequence (FASTA) and gene annotation (GTF) files are now split and formatted for Cell Ranger ARC, with file paths specified in the .config file, and Cell Ranger ARC can now be run.

#### Prepare `config` File

``` bash
cat 03_cellranger_arc_barley.config
```

``` json
{
    "organism": "barley",
    "genome": ["MOREX3_subset_v3"],
    "input_fasta": ["/data/morex3_split_FINAL_1H_2H.fasta"],
    "input_gtf": ["/extdata/B1_FINAL_MODIFIED_GTF_subset_1H_2H_morex_3.gtf"],
}
```

> **Note:** In the full genome (not the subset), you need to add infomation about mitochondrial and chloroplast to the "non_nuclear_contigs" field in the config file: `"non_nuclear_contigs": ["Mt-0-525599", "Pt-0-115974"]` in the config file

#### Run Cell Ranger ARC

``` bash
CONFIG_IN="03_cellranger_arc_barley.config"

cellranger-arc mkref \
  --config=${CONFIG_IN} \
  --nthreads=24 \
  >mkref.log 2>&1
```

## Citation

Please cite this tool if you use it

```{r citation, eval=TRUE}
citation("scShardSplitRef")
```

## License

This project is licensed under the [GPL-3.0 license](LICENSE.md).

## Contributing

Please refer to the [CONTRIBUTING.md]() \<-- TODO: Provide link once this is public ---!\>
If you'd like to help improve this tool, please send your suggestions via email.
Please clearly outline the feature(s) you think should be added.
As the tool develops, new features can be added as separate package functions and contributed back to the project.\
Remember that improving tools is important, but keeping them neat and not over-complicated is just as essential.

## Acknowledgements

This work was supported by resources provided by the [Pawsey Supercomputing Research Centres](https://pawsey.org.au/) Nimbus Research Cloud (<https://doi.org/10.48569/v0j3-qd51>), with funding from the Australian Government and the Government of Western Australia.

GRDC funded this project through the Analytics for the Australian Grains Industry (AAGI) investment, CUR2210-005OPX (AAGI-CU).
