---
title: "scShardSplitRef"
author: "Irina Kuznetsova"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scShardSplitRef}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
devtools::load_all()  #### delete after publishing
```


## Package overview

{scShardSplitRef} designed for researchers working with multiome single-cell data from species that require building a custom reference genome, particularly when one or more chromosomes/contigs in the reference FASTA exceed the Cell Ranger ARC limit of 536.8Mb (2\^29 bases).
{scShardSplitRef} enables splitting of long chromosomes/contigs generating shorter fragments that are compitable with Cell Ranger ARC requirements.
The software outputs FASTA and GTF files containing the split chromosomes/contigs, which can be directly processed by `cellranger-arc mkref`.

To familiarise yourself with the package and workflow of preparing reference genome sequence (FASTA file) and gene annotations (GTF file) for Cell Ranger ARC we suggest starting with a guided walk through.
We are using barley genome (MOREX V3) as an example genome.
Example data files are distributed with the package for example purposes.

## {scShardSplitRef} guided walk through

### Installation

```{r setup}
library("scShardSplitRef")
library("knitr") # used for a kable table later in the document
```

### Prerequisites


| Software        | Version | Purpose                         |
|-----------------|---------|---------------------------------|
| bedtools        | ≥ 2.30  | BED file operations             |
| Cell Ranger ARC | ≥ 2.0   | Single-cell multiome processing |


### Input files


| File                   | Format   | Description                             |
|------------------|---------------------|---------------------------------|
| Genome sequence        | FASTA    | Reference genome sequence               |
| Gene annotations       | GTF      | Gene features and coordinates           |
| Centromere coordinates | BED-like | Centromere positions per chromosome (x) |


> (x) **Note:** If centromere coordinates are not available for your species, you can manually create a BED-like file with the midpoint position of each chromosome.
> {scShardSplitRef} will then check for any overlaps with genic regions and generate the correct BED files for splitting.

For this tutorial, we will be using a genome sequence (FASTA) and gene annotations (GTF) of the barley genome (*Hordeum Vulgare, Morex V3*) available from [Ensembl](https://ftp.ensemblgenomes..ebi.ac.uk) and centromere coordinates from [GrainGenes](https://graingenes.org/GG3/content/morex-v3-files-2021).

<details>

<summary><b>What does a GTF file for MOREX V3 look like?</b></summary>

The format of the GTF file is a tab-delimited 9-column file.

```{r}
# Gene annotations (GTF) file:
# data.dir = "/inst/extdata/01_Hordeum_vulgare.Mt_Pt_v49_Barley_v62_COMBINED.gtf"

gtf_file <- system.file(
  "extdata/01_Hordeum_vulgare.Mt_Pt_v49_Barley_v62_COMBINED.gtf",
  package = "scShardSplitRef"
)
gtf_df <- read.table(file = gtf_file, header = FALSE, sep = "\t")
dim(gtf_df)
gtf_df[1:10, ]
```

</details>

The format of the centromere coordinates file is a tab-delimited 3-column file, where the 1st column contains chromosome/contig names, the 2nd column contains the centromere coordinate, and the 3rd column contains chromosome/contig region end coordinates.

<details>

<summary><b>What does BED-like centromere files for MOREX V3 look like?</b></summary>

```{r}
# Manually formatted centromere file (please check what it should look like below and check **Step-0** description):
# data.dir = "/inst/extdata/02_MorexV3_centromere_coordinates.bed"

bed_file <- system.file(
  "extdata/02_MorexV3_centromere_coordinates.bed",
  package = "scShardSplitRef"
)
centr_bed_df <- read.table(file = bed_file, header = FALSE, sep = "\t")
dim(centr_bed_df) ## 292 x 3
centr_bed_df[1:10, ]
```

</details>

## Step-by-step Guide

We begin by assessing the formatting of the input GTF and BED files provided by the user, followed by checking and, if necessary, modifying the split coordinates in the BED file.
The GTF file is then updated according to the split position.
Finally, the FASTA file is split and formatted for compatibility with Cell Ranger ARC.
All files need to be decompressed for use in CellRanger.

### Step-1

The first function called `validate_inputs()`, checks the correct format of (FASTA and BED-like) input files.
The centromere coordinate file is manually created in BED-like format and should be tab-delimited.
All files need to be decompressed for use in CellRanger.

#### The help page for the validate_inputs() 

```{r, eval=FALSE}
?validate_inputs
```


####  Validate (FASTA and BED) input files

```{r}
# Load MOREX V3 files
# Note1: all files need to be decompressed for use in CellRanger
# Note2: use correctly formatted BED file with centromere coordinates
bed_path <- system.file(
  "extdata/02_MorexV3_centromere_coordinates.bed",
  package = "scShardSplitRef"
)
gtf_path <- system.file(
  "extdata/01_Hordeum_vulgare.Mt_Pt_v49_Barley_v62_COMBINED.gtf",
  package = "scShardSplitRef"
)
# Validate input (FASTA and BED) files
input_verif <- validate_inputs(bed = bed_path, gtf = gtf_path)
input_verif
```

> If GTF and BED formats are correct, a confirmation message will be printed to the console:\ *"All files loaded and validated successfully"*

### Step-2

The next step is to prepare split coordinates in appropriate BED-like format.
This function ensures that chromosomes are not split in the middle of a gene region.
To achieve this, we first check whether the centromere coordinates overlap with any gene features, and if they do, we shift the split position by a specified length in base pairs (bp).

### The help page for the determine_split_regions() function 

```{r, eval=FALSE}
?determine_split_regions
```

#### Define output directory

```{r}
# Build BED-like split intervals from genomic regions (BED) and gene annotations (GTF)

# Define output directory
output_file <- file.path(tempdir(), "outdata")
dir.create(output_file, recursive = TRUE, showWarnings = FALSE)
output_file_name_dir <- file.path(output_file, "B1_Split_regions.bed")

# Determine split regions
get_split_reg <- determine_split_regions(
  bed = bed_path,
  gtf = gtf_path,
  output_bed = output_file_name_dir,
  limit = 2L^29L,
  shift_by = 20L
)

get_split_reg
```

<details>

<summary><b>What does the generated BED file look like?</b></summary>

```{r}
bed_f <- system.file(
  "outdata/B1_Split_regions.bed",
  package = "scShardSplitRef"
)
split_region_bed_df <- read.table(file = bed_f, header = FALSE, sep = "\t")
dim(split_region_bed_df)
split_region_bed_df[1:10, ]
```

</details>

### Step-3

Generate a gene annotation (GTF) file in compatible to Cell Ranger ARC format.
The first column of the GTF file should be in the following form `Chr-RegionStart-RegionEnd`.
Note that the output file name is hardcoded and starts with "B1_FINAL_MODIFIED_GTF\_[*genome_name*]\_[*genome_version*].gtf".

#### The help page for the `process_gtf()` function

```{r, eval=FALSE}
?process_gtf
```

#### Generate gene annotation (GTF) file in compatible to Cell Ranger ARC format 

```{r}
# Define output directory
gtf_path <- system.file(
  "extdata/01_Hordeum_vulgare.Mt_Pt_v49_Barley_v62_COMBINED.gtf",
  package = "scShardSplitRef"
)

# Process GTF file
GTF_processed <- process_gtf(
  split_regions_bed = output_file_name_dir,
  gtf = gtf_path,
  genome_name = "barley",
  genome_version = "3",
  out_path = "."
)

GTF_processed
```

> The output file is in the GTF format.
> Note that the first column has been reformatted — the regions are now split.
> The chromosome name is formatted as `1H-0-206486643`, which is the region from the beginning of the chromosome to the split position; \``1H-206486643-516505932` is the region from the split position to the end of the chromosome.

<details>

<summary><b>What does the generated GTF file look like?</b></summary>

```{r}
gtf_f <- system.file(
  "outdata/B1_FINAL_MODIFIED_GTF_barley_3.gtf",
  package = "scShardSplitRef"
)
processed_gtf <- read.table(file = gtf_f, header = FALSE, sep = "\t")
dim(processed_gtf)

sep <- as.data.frame(matrix("...", nrow = 1, ncol = 5))
colnames(sep) <- colnames(processed_gtf)[1:5]

kable(rbind(
  processed_gtf[200:203, 1:5],
  sep,
  processed_gtf[(nrow(processed_gtf) - 2):nrow(processed_gtf), 1:5]
))
```

</details>

### Step-4

Final steps is to split the FASTA file, where the header lines should have the same chrosomome annotations as the GTF file.

```{bash, eval=FALSE}
# Please note that due to file size limitations, the barley genome FASTA file is not available on our GitHub repository at /inst/extdata/
IN_FA = "/inst/extdata/01_Hordeum_vulgare.MorexV3_pseudomolecules_assembly.dna.primary_assembly_COMBINED_Mt_Pt_v49_Barley_v62.fa"
BED_FIN_OUT = "/inst/outdata/B1_Split_regions.bed"
FASTA_FIN_OUT = "/inst/outdata/"

# Split FASTA
bedtools getfasta -fi ${IN_FA} -bed ${BED_FIN_OUT} -fo ${FASTA_FIN_OUT}/B1_FASTA_split.fasta

# Convert FASTA heading in Cell Ranger ARC compatible format: Chr-RegionStart-RegionEnd
sed '/^>/s/:/-/g' ${FASTA_FIN_OUT}/B1_FASTA_split.fasta >${FASTA_FIN_OUT}/B1_FASTA_split_final.fasta
```

<details>

<summary><b>What does the generated FASTA file look like?</b></summary>

```
# B1_FASTA_split_final.fasta

>1H:0-206486643
GCATTTGGTTGCTTTTATTCTAAACTA...
>1H:206486643-516505932
ATCAGTACATGAGCTAGTATTAGAGGCTAGACTAGGGATCAATTGGTGTTTGTA...
>2H:0-301293086
GCCCTGAC...
...
>CAJHDD010000283.1:0-50002
TATTTGCCACTTCAATCTGAT..
>Mt:0-525599
TGAGAGAAAAAATAAGACTCTTTTAAGATTGCAAAACGTTTTGCAATGAA..
>Pt:0-115974
TCCCACTAAACATGT...

```

</details>

### Step-5

The genome sequence (FASTA) and gene annotation (GTF) files are now split and formatted for Cell Ranger ARC.
The file paths are specified in the `.config` file and Cell Ranger ARC can now be run.

#### Prepare config file

```{bash, eval=FALSE}
cat S0_cellranger_arc_barley.config
```

``` json
{
    "organism": "barley",
    "genome": ["MOREX3_v3"],
    "input_fasta": ["B1_FASTA_split_final.fasta"],
    "input_gtf": ["/outdata/B1_FINAL_MODIFIED_GTF_barley_3.gtf"],
    "non_nuclear_contigs": ["Mt-0-525599", "Pt-0-115974"]
}
```

#### Run Cell Ranger ARC

```{bash, eval=FALSE}
CONFIG_IN="S0_cellranger_arc_barley.config"

cellranger-arc mkref \
  --config=${CONFIG_IN} \
  --nthreads=24 \
  >mkref.log 2>&1
```

## R package session information

<details>

<summary><b>sessionInfo()</b></summary>

```{r sessioninfo}
sessionInfo()
```

</details>

## Citation

Please cite this tool if you use it

```{r}
library("scShardSplitRef")
citation("scShardSplitRef")
#> Warning in citation("scShardSplitRef"): could not determine year for 'scShardSplitRef'
#> from package DESCRIPTION file
#> To cite package 'scShardSplitRef' in publications use:
#> 
#>   Kuznetsova I, Pembleton L, Curci L, Sparks A (????). _scShardSplitRef: Build
#>   Custom Reference Genome for Single-Cell Multiome ATAC and Gene Expression Data
#>   of Large-genome Species._. R package version 0.0.0.9000.
#> 
#> A BibTeX entry for LaTeX users is
#> 
#>   @Manual{,
#>     title = {scShardSplitRef: Build Custom Reference Genome for Single-Cell Multiome ATAC and
#> Gene Expression Data of Large-genome Species.},
#>     author = {Irina Kuznetsova and Luke Pembleton and Luca Curci and Adam H. Sparks},
#>     note = {R package version 0.0.0.9000},
#>   }
```

## License

This project is licensed under the [GPL-3.0 license](LICENSE.md).

## Contributing

If you'd like to help improve this tool, please send your suggestions via email.\
Please clearly outline the feature(s) you think should be added. As the tool develops, new features can be added as separate package functions and contributed back to the project.\
Remember that improving tools is important, but keeping them neat and not over-complicated is just as essential.

## Acknowledgements

This work was supported by resources provided by the [Pawsey Supercomputing Research Centres](https://pawsey.org.au/) Nimbus Research Cloud (<https://doi.org/10.48569/v0j3-qd51>), with funding from the Australian Government and the Government of Western Australia.

GRDC (?)
