---
title: "scShardSplitRef"
author: "Irina Kuznetsova"
output: rmarkdown::html_vignette    ####html_vignette
vignette: >
  %\VignetteIndexEntry{scShardSplitRef}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  )
devtools::load_all() #### delete after publishing
```

## 1. Purpose

This toy dataset provides a minimal example to demonstrate the logic of the algorithm that prepares reference files (GTF and FASTA) for ```cellranger‑arc mkref```. It allows users to test the workflow without needing large genome files and helps confirm that the proposed workflow is functioning correctly. It also provides an example of how to use it.

## 2. {scShardSplitRef} guided walk through using a toy example

### Installation

```{r setup}
library("scShardSplitRef")
library("knitr") # used for a kable table later in the document
```

### Prerequisites

| Software | Version | Purpose             |
|----------|---------|---------------------|
| bedtools | ≥ 2.30  | BED file operations |

### Input files

| File | Format | Description | File Name |
|----------------|----------------|--------------------|--------------------|
| Gene annotations | GTF | Gene features and coordinates | extdata/A3_toy_all_scenarios_2chr.gtf |
| Centromere coordinates | BED-like | Centromere positions per chromosome (x) | extdata/A3_toy_centromeres.bed |
| Genome sequence | FASTA | Reference genome sequence | extdata/A3_toy.fasta |

<br>


<details>
<summary><b>What does a GTF file for MOREX V3 look like?</b></summary>

```{r}
# Gene annotations (GTF) file. The format of the GTF file is a tab-delimited 9-column file.
# data.dir = "/inst/extdata/01_Hordeum_vulgare.Mt_Pt_v49_Barley_v62_COMBINED.gtf"

gtf_file <- system.file(
  "extdata/A3_toy_all_scenarios_2chr.gtf",
  package = "scShardSplitRef"
  )
gtf_df <- read.table(file = gtf_file, 
                     header = FALSE, 
                     sep = "\t"
                     )
dim(gtf_df)
gtf_df[1:10, ]
```

</details>



<details>

<summary><b>What does a BED-like centromere file for MOREX V3 look like?</b></summary>

```{r}
# The format of the centromere coordinates file is a tab-delimited 3-column file, where the 1st column contains chromosome/contig names, the 2nd column contains the centromere coordinate, and the 3rd column contains chromosome/contig region end coordinates.
# Manually formatted centromere file (please check what it should look like below):
# data.dir = "/inst/extdata/02_MorexV3_centromere_coordinates.bed"

bed_file <- system.file(
  "extdata/02_MorexV3_centromere_coordinates.bed",
  package = "scShardSplitRef"
  )
centr_bed_df <- read.table(file = bed_file, 
                           header = FALSE, 
                           sep = "\t"
                           )
dim(centr_bed_df) ## 292 x 3
centr_bed_df[1:10, ]
```

</details>

## Step-by-step Guide

We begin by assessing the formatting of the input GTF and BED files, followed by checking and, if necessary, modifying the split coordinates in the BED file.
The GTF file is then updated according to the split position.
Finally, the FASTA file is split and formatted for compatibility with Cell Ranger ARC.


### Step-1

The first function called `validate_inputs()`, checks the correct format of (FASTA and BED-like) input files.
The centromere coordinate file is manually created in BED-like format and should be tab-delimited.
All files need to be decompressed for use in Cell Ranger.

#### The help page for the validate_inputs() 

```{r, eval=FALSE}
?validate_inputs
```


####  Validate (FASTA and BED) input files

```{r}
bed_path <- system.file(
  "extdata/A3_toy_centromeres.bed",
  package = "scShardSplitRef"
  )
gtf_path <- system.file(
  "extdata/A3_toy_all_scenarios_2chr.gtf",
  package = "scShardSplitRef"
  )
# Validate input (FASTA and BED) files
input_verif <- validate_inputs(bed = bed_path, 
                               gtf = gtf_path
                               )
kable(input_verif$bed[1:5,])
kable(input_verif$gtf[1:5,])
```

> If GTF and BED formats are correct, a confirmation message will be printed to the console:\ *"All files loaded and validated successfully"*

### Step-2
The next step is to prepare split coordinates in appropriate BED-like format.
This function ensures that chromosomes are not split in the middle of a gene region.
To achieve this, we first check whether the centromere coordinates overlap with any gene features, and if they do, we shift the split position by a specified length in base pairs (bp).

### The help page for the determine_split_regions() function 

```{r, eval=FALSE}
?determine_split_regions
```

#### Define output directory

```{r}
# Build BED-like split intervals from genomic regions (BED) and gene annotations (GTF)

# Define output directory
output_file <- file.path(tempdir(), 
                         "outdata"
                         )
dir.create(output_file, 
           recursive = TRUE, 
           showWarnings = FALSE
           )
output_file_name_dir <- file.path(output_file, 
                                  "A3_toy_split_regions.bed"
                                  )

# Determine split regions
get_split_reg <- determine_split_regions(
  bed = bed_path,
  gtf = gtf_path,
  output_bed = output_file_name_dir,
  limit = 2L^29L,
  shift_by = 20L
  )

get_split_reg
```

<details>

<summary><b>What does the generated BED file look like?</b></summary>

```{r}
bed_f <- system.file(
  "outdata/A3_toy_split_regions.bed",
  package = "scShardSplitRef"
)
split_region_bed_df <- read.table(file = bed_f, 
                                  header = FALSE, 
                                  sep = "\t"
                                  )
dim(split_region_bed_df)
split_region_bed_df[1:10, ]
```

</details>

### Step-3

Generate a gene annotation (GTF) file in compatible to Cell Ranger ARC format.
The first column of the GTF file should be in the following form `Chr-RegionStart-RegionEnd`.
Note that the output file name is hardcoded and starts with "B1_FINAL_MODIFIED_GTF\_[*genome_name*]\_[*genome_version*].gtf".

#### The help page for the `process_gtf()` function

```{r, eval=FALSE}
?process_gtf
```

#### Generate gene annotation (GTF) file in compatible to Cell Ranger ARC format 

```{r}
# Define output directory
gtf_path <- system.file(
  "extdata/A3_toy_all_scenarios_2chr.gtf",
  package = "scShardSplitRef"
)

# Process GTF file and write file to tempdir()
process_gtf(
  split_regions_bed = output_file_name_dir,
  gtf = gtf_path,
  genome_name = "toy",
  genome_version = "noversion",
  out_path = tempdir()
  )
```

> The output file is in the GTF format.
> Note that the first column has been reformatted — the regions are now split.
> The chromosome name is formatted as `1H-0-206486643`, which is the region from the beginning of the chromosome to the split position; \``1H-206486643-516505932` is the region from the split position to the end of the chromosome.

<details>

<summary><b>What does the generated GTF file look like?</b></summary>

```{r}
gtf_f <- system.file(
  "outdata/B1_FINAL_MODIFIED_GTF_toy_noversion.gtf",
  package = "scShardSplitRef"
  )

processed_gtf <- read.table(file = gtf_f, 
                            header = FALSE, 
                            sep = "\t"
                            )
dim(processed_gtf)

sep <- as.data.frame(matrix("...", 
                            nrow = 1, 
                            ncol = 5)
                     )
colnames(sep) <- colnames(processed_gtf)[1:5]

kable(rbind(
  processed_gtf[10:23, 1:5],
  sep,
  processed_gtf[(nrow(processed_gtf) - 2):nrow(processed_gtf), 1:5]
  )
  )
```

</details>

### Step-4

Final steps is to split the FASTA file, where the header lines should have the same chrosomome annotations as the GTF file.

```{bash, eval=FALSE}
# Please note that due to file size limitations, the barley genome FASTA file is not available on our GitHub repository at /inst/extdata/
IN_FA="/inst/extdata/A3_toy.fasta"
BED_FIN_OUT="/inst/outdata/A3_toy_centromeres.bed"
FASTA_FIN_OUT="/inst/outdata/"

# Split FASTA
bedtools getfasta -fi ${IN_FA} -bed ${BED_FIN_OUT} -fo ${FASTA_FIN_OUT}/A3_toy_bedtools.fasta
```

<details>

<summary><b>What does the generated FASTA file look like?</b></summary>

```
# outdata/A3_toy_bedtools.fasta

>1H:0-160
ATCGGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAATCGGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTA
>1H:160-200
GCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTA
>2H:0-180
GCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTA
>2H:180-300
GCTAGCTAGCTAGCTAGCATCGGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGC
>CAJHDD010000004.1:0-50
ATCGGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGC
>CAJHDD010000001.1:0-40
GCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTAGCTA
>Mt:0-20
ATCGGCTAGCTAGCTAGCTA
>Pt:0-10
ATCGGCTAGC

```

</details>




### Output Files 
Output files to be used with Cell Ranger ARC.


| File | Format | Description | File Name |
|----------------|----------------|--------------------|--------------------|
| Gene annotations | GTF | Gene features and coordinates | outdata/B1_FINAL_MODIFIED_GTF_toy_noversion.gtf |
| Genome sequence | FASTA | Reference genome sequence | outdata/A3_toy_bedtools.fasta |



## 3. Session Info

```{r session-info}
sessionInfo()
```

## 4. Acknowledgements

This work was supported by resources provided by the Pawsey Supercomputing Research Centres Nimbus Research Cloud (<https://doi.org/10.48569/v0j3-qd51>), with funding from the Australian Government and the Government of Western Australia.
