---
title: "scShardSplitRef"
author: "Irina Kuznetsova"
output: rmarkdown::md_document    ####html_vignette
vignette: >
  %\VignetteIndexEntry{scShardSplitRef}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



## 1. Package overview  

*scShardSplitRef* designed for researchers working with single-cell data from species that require building a custom reference genome, particularly when one or more contigs in the reference FASTA exceed the Cell Ranger limit of 536.8Mb (2^29 bases). *scShardSplitRef* automates this splitting process, generating contig fragments that comply with Cell Ranger requirements. The software outputs FASTA and GTF files containing the split contigs, which can be directly processed by `cellranger-arc mkref`. 

This real-world example demonstrates a step-by-step workflow for downloading and preparing barley (MOREXv3) reference files (GTF and FASTA) for `cellranger-arc mkref`. 




## 2. Installation  
Start by loading required packages:  

```{r}
library("dplyr"); library("stringr"); library("utils")
``` 



## 3. Accessing Package Help and Vignettes  

#library(scShardSplitRef)
#help(package="scShardSplitRef")
#vignette(package="scShardSplitRef")
#vignette(package="scShardSplitRefQuickStart", "scShardSplitRefDetailed")
#?scShardSplitRef
  


## 4. Quick start

## **RUN THE ANALYSIS BY CALLING THE FUNCTIONS**
##### 1) Check if gene features are overlapping with the centromere position  

```
gtf_path_T="/tests/testthat/testdata/A3_toy_all_scenarios_2chr.gtf"
centromere_path_T="/tests/testthat/testdata/IN0_toy_centromeres_for_gtf.bed"
out_path_T="/tests/testthat/testdata/testoutput/"
centromere_split = prepare_centromere_split(
    gtf_path_T, 
    centromere_path_T, 
    out_path_T
    )
```

######## Output message:

```
No overlapping gene features detected ? use the original centromere coordinates BED file.
Use:
tests/testthat/testdata/IN0_toy_centromeres_for_gtf.bed as the input to modify GTF file.
and:
/tests/testthat/testdata/testoutput/IN0_original_centromeres_for_FASTA.bed for splitting FASTA
```

######## Depending on the input data, the message could be:  

```Overlapping gene features with centromere coordinates detected!
re-calculating split coordinates for these chromosomes
Use:
/tests/testthat/testdata/testoutput/RECAL1_FINAL_SPLIT_coord_for_GTF.bed as the input to modify GTF file.
and:
/tests/testthat/testdata/testoutput/RECAL1_FINAL_SPLIT_coord_for_FASTA.bed
 as the input for splitting FASTA.
```


##### 2) Validate inputs 
```
centromere_path2="/tests/testthat/testdata/testoutput/RECAL1_FINAL_SPLIT_coord_for_GTF.bed" ### Depending on the input data, the input might be tests/testthat/testdata/IN0_toy_centromeres_for_gtf.bedtests/testthat/testdata/IN0_toy_centromeres_for_gtf.bed
validate_inputs(
    gtf_path, 
    centromere_path2)
```

#### 3) Process GTF
```
unchar_region="CAJHDD"
mito="Mt"
chloro="Pt"
processed_GTF_First_Fith_Sixth_Col = process_gtf(
    gtf_path, 
    centromere_path2, 
    unchar_region, 
    mito, 
    chloro
    )
```

#### 4) Process 9th column of the GTF file
```
keep_attributes= c("gene_id", "transcript_id", "gene_name")
out_path="testthat/testdata/testoutput/"
genome_name ="BarleyMOREXV3_toy"
genome_version ="v51_ENSEMBL"
final_result = process_attributes_column(
    processed_GTF_First_Fith_Sixth_Col, 
    keep_attributes,
    out_path,
    genome_name,
    genome_version
    )
```  







## 5. Session Info  

```{r session-info}
sessionInfo()
```



## 6. Acknowledgements  
This work was supported by resources provided by the [Pawsey Supercomputing Research Centres](https://pawsey.org.au/) Nimbus Research Cloud (https://doi.org/10.48569/v0j3-qd51), with funding from the Australian Government and the Government of Western Australia.


## not used , will be removed
#Start by installing the BiocManager package:  
#```{r}
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("scShardSplitRef")
#```  