---
title: "scShardSplitRef"
author: "Irina Kuznetsova"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scShardSplitRef}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Package overview

*scShardSplitRef* designed for researchers working with single-cell data from species that require building a custom reference genome, particularly when one or more contigs in the reference FASTA exceed the Cell Ranger ARC limit of 536.8Mb (2\^29 bases). *scShardSplitRef* automates this splitting process, generating contig fragments that comply with Cell Ranger requirements. The software outputs FASTA and GTF files containing the split contigs, which can be directly processed by `cellranger-arc mkref`.

This real-world example demonstrates a step-by-step workflow for downloading and preparing barley (MOREXv3) reference files (GTF and FASTA) for `cellranger-arc mkref`.

To familiarise yourself with the package and workflow of preparing reference genome sequence (FASTA file) and gene annotations (GTF file) to be compatible with Cell Ranger ARC we suggest starting with a guided walk through that uses MOREXv3 available from our GitHub page.

To get familiar with the package and its workflow, we recommend starting with a guided walkthrough. It covers how to prepare a reference genome sequence (FASTA file) and gene annotations (GTF file) that are compatible with Cell Ranger ARC, using MOREXv3 as an example genome. available from our GitHub page.



## 2. Accessing Package Help and Vignettes

```{r, eval=TRUE, setup}
library(scShardSplitRef)
help(package = "scShardSplitRef")

vignette(package = "scShardSplitRef")

help("scShardSplitRef")
```

## 4. Quick start

### 1) Check if gene features are overlapping with the centromere position

```{r, check_gene_features}
 regions <- system.file(
   "extdata",
   "IN0_toy_centromeres_for_gtf.bed",
   package = "scShardSplitRef",
   mustWork = TRUE
 )
 genes <- system.file(
   "extdata",
   "A3_toy_all_scenarios_2chr.gtf",
   package = "scShardSplitRef",
   mustWork = TRUE
 )

 determine_split_regions(
   regions_bed = regions,
   genes_gtf = genes,
   output_bed = file.path(tempdir(), "split_regions.bed"),
   limit = 2L^29L,
   shift_by = 1L
 )
```

### 2) Validate inputs

```{r, validate_inputs}
gtf_path <- system.file(
  "extdata",
  "A3_toy_all_scenarios_2chr.gtf",
  package = "scShardSplitRef",
  mustWork = TRUE
)

centromere_path2 <- system.file(
  "extdata",
  "RECAL1_FINAL_SPLIT_coord_for_GTF.bed",
  package = "scShardSplitRef",
  mustWork = TRUE
)

validate_inputs(
  gtf_path,
  centromere_path2
)
```

### 3) Process GTF

```{r, process_gtf}
processed_GTF_First_Fith_Sixth_Col <- process_gtf(
  gtf_path,
  centromere_path2
)

processed_GTF_First_Fith_Sixth_Col
```

### 4) Process 9th column of the GTF file

```{r, process_col_9}
keep_attributes <- c("gene_id", "transcript_id", "gene_name")
out_path <- tempdir() # write only to R's tempdirectory
genome_name <- "BarleyMOREXV3_toy"
genome_version <- "v51_ENSEMBL"

final_result <- process_attributes_column(
  result = processed_GTF_First_Fith_Sixth_Col,
  keep_attributes,
  out_path,
  genome_name,
  genome_version
)

final_result
```

## Acknowledgements

This work was supported by resources provided by the [Pawsey Supercomputing Research Centre's](https://pawsey.org.au/) Nimbus Research Cloud (<https://doi.org/10.48569/v0j3-qd51>), with funding from the Australian Government and the Government of Western Australia.

GRDC AAGI project support code: AAGI–CU–RD–SNB #TODO: Is there any standardised GRD message for acknowledgment?
